### 2025-04-05 decision-making.qmd
20250405 - wisdom of the crowd

Here is a code example of when the crowd-averaged predictions are centered on the truth for a player.
The distribution of projected points versus truth is depicted in @fig-bracketingCentered.

@@

@@

As demonstrated above, when the crowd-averaged predictions are centered on the truth, the crowd-averaged predictions are perfectly accurate and are more accurate (MAE/MdAE = `{r} maeBracketingCentered_crowd`) than the accuracy of the typical individual forecaster (MAE = `{r} maeBracketingCentered_individual`; MdAE = `{r} mdaeBracketingCentered_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError) and [median absolute error](#sec-medianAbsoluteError).

Here is a code example of when the crowd-averaged predictions bracket but are not centered on the truth for a player, with a high bracketing rate.
The distribution of projected points versus truth is depicted in @fig-bracketingNotCentered1.

@@

@@

As demonstrated above, when the crowd-averaged predictions bracket but are not centered on the truth, and the bracketing rate is high, the crowd-averaged predictions (MAE/MdAE = `{r} maeBracketingNotCentered1_crowd`) are more accurate than the accuracy of the typical individual forecaster (MAE = `{r} maeBracketingNotCentered1_individual`; MdAE = `{r} mdaeBracketingNotCentered1_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError) and [median absolute error](#sec-medianAbsoluteError).

Here is another code example of when the crowd-averaged predictions bracket but are not centered on the truth for a player, with a low bracketing rate.
The distribution of projected points versus truth is depicted in @fig-bracketingNotCentered2.

```{r}
predictedValues <- c(10,29,29,40,60,70,80,90)
actualValue <- 15
errors <- predictedValues - actualValue

@@

```{r}
#| label: fig-bracketingNotCentered2
#| fig-cap: "Example Distribution of Projected Fantasy Points that Brackets But is Not Centered on the Truth."
#| fig-alt: "Example Distribution of Projected Fantasy Points that Brackets But is Not Centered on the Truth."
#| code-fold: true

bracketingNotCentered2 <- petersenlab::wisdomOfCrowd(
  predicted = predictedValues,
  actual = actualValue
)

maeBracketingNotCentered2_individual <- bracketingNotCentered2[which(row.names(bracketingNotCentered2) == "individual"), "MAE"]
maeBracketingNotCentered2_crowd <- bracketingNotCentered2[which(row.names(bracketingNotCentered2) == "crowdAveraged"), "MAE"]

mdaeBracketingNotCentered2_individual <- bracketingNotCentered2[which(row.names(bracketingNotCentered2) == "individual"), "MdAE"]
meBracketingNotCentered2_crowd <- bracketingNotCentered2[which(row.names(bracketingNotCentered2) == "crowdAveraged"), "ME"]

ggplot2::ggplot(,
  mapping = aes(
    x = predictedValues)
) +
  geom_density(
    fill = "lightgray"
  ) +
  geom_vline(
    xintercept = mean(predictedValues),
    color = "darkgray",
    linewidth = 1.5) +
    geom_vline(
    xintercept = actualValue,
    linetype = "dashed",
    linewidth = 1.5
  ) +
  annotate(
    "segment",
    x = mean(predictedValues),
    xend = actualValue,
    y = 0,
    yend = 0,
    linewidth = 1.5,
    arrow = arrow(
      angle = 20,
      ends = "both",
      type = "closed")
  ) +
  annotate(
    "text",
    x = actualValue + (meBracketingNotCentered2_crowd / 2),
    y = 0.0003,
    label = paste("Mean Error: ", meBracketingNotCentered2_crowd, sep = ""),
    hjust = 0.5 # center-justify
  ) +
  labs(
    x = "Fantasy Points",
    y = "Density",
    title = "Density Plot of Predicted Fantasy Points for a Player",
    subtitle = "Crowd-Averaged Projection (Vertical Gray Line) Vs.\nTruth, i.e., Actual Fantasy Points (Dashed Line)"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title
```

As demonstrated above, when the crowd-averaged predictions bracket but are not centered on the truth and the bracketing rate is low, the crowd-averaged predictions (MAE/MdAE = `{r} maeBracketingNotCentered2_crowd`) are more accurate than the accuracy of the typical individual forecaster (MAE = `{r} maeBracketingNotCentered2_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError), but they can still be less accurate—in terms of [median absolute error](#sec-medianAbsoluteError)—than the individual forecaster who is at the 50th percentile in accuracy (MdAE = `{r} mdaeBracketingNotCentered2_individual`).

Here is a code example of when the crowd-averaged predictions do not bracket the truth for a player, but the distribution is still close to truth.
The distribution of projected points versus truth is depicted in @fig-noBracketing1.

@@

As demonstrated above, when the crowd-averaged predictions do not bracket the truth but are still close to the truth, the crowd-averaged predictions (MAE/MdAE = `{r} maeNoBracketing1_crowd`) are as accurate as the accuracy of the typical individual forecaster (MAE = `{r} maeNoBracketing1_individual`; MdAE = `{r} mdaeNoBracketing1_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError) and [median absolute error](#sec-medianAbsoluteError).

@@

```{r}
predictedValues <- c(0,65,70,75,80,85,90,95)
actualValue <- 100
errors <- predictedValues - actualValue

@@

In conclusion, as demonstrated from these examples, when the crowd-averaged predictions are centered on the truth, the crowd-averaged predictions are perfectly accurate and are more accurate than the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError) and of most individual forecasters in terms of [median absolute error](#sec-medianAbsoluteError).
When the crowd-averaged predictions bracket but are not centered on the truth, the crowd-averaged predictions are more accurate than the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError).
If the bracketing rate is high, the crowd-averaged predictions will be more accurate—in terms of [median absolute error](#sec-medianAbsoluteError)—than most individual forecasters; however, if the bracketing rate is low, the crowd-averaged predictions can be less accurate than than most individual forecasters.
When the crowd-averaged predictions do not bracket the truth, the crowd-averaged predictions are as accurate as the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError).
Depending on how far some of the individual projections are from the truth, the crowd-averaged projections can be less accurate—in terms of [median absolute error](#sec-medianAbsoluteError)—than most individual forecasters.

In sum, going with crowd-averaged projections will always yield predictions that are as or more accurate (in terms of [mean absolute error](#sec-meanAbsoluteError)) than the (collective) accuracy of the individuals' projections.
However, the crowd-averaged projections may be less accurate than any *individual* projection, including the projection of the forecaster who is at the 50th percentile in accuracy.
That is, depending on the bracketing rate, the crowd-averaged projections can still be less accurate than most individual projections.
In general, the higher the bracketing rate and the more closely that the projections are centered on the truth, the more accurate the crowd-averaged projections will be.

>>>>

Here is a code example of when the individual predictions (from which the crowd-averaged predictions are derived) are centered on the truth for a player.
The distribution of projected points versus truth is depicted in @fig-bracketingCentered.

@@

bracketingCentered_bracketingRate <- bracketingCentered[which(row.names(bracketingCentered) == "crowdAveraged"), "bracketingRate"] * 100

@@

As demonstrated above, when the individual predictions are centered on the truth, the bracketing rate is very high (in this case `{r} petersenlab::apa(bracketingCentered_bracketingRate, decimals = 2)`%), and the crowd-averaged predictions are perfectly accurate and are more accurate (MAE/MdAE = `{r} maeBracketingCentered_crowd`) than the accuracy of the typical individual forecaster (MAE = `{r} maeBracketingCentered_individual`; MdAE = `{r} mdaeBracketingCentered_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError) and [median absolute error](#sec-medianAbsoluteError).

Here is a code example of when the individual predictions bracket but are not centered on the truth for a player, with a high bracketing rate and no strong outliers.
The distribution of projected points versus truth is depicted in @fig-bracketingNotCentered1.

@@

bracketingNotCentered1_bracketingRate <- bracketingNotCentered1[which(row.names(bracketingNotCentered1) == "crowdAveraged"), "bracketingRate"] * 100

@@

As demonstrated above, when the individual predictions bracket but are not centered on the truth, and the bracketing rate is high (in this case `{r} petersenlab::apa(bracketingNotCentered1_bracketingRate, decimals = 2)`%) with no strong outliers, the crowd-averaged predictions (MAE/MdAE = `{r} maeBracketingNotCentered1_crowd`) are more accurate than the accuracy of the typical individual forecaster (MAE = `{r} maeBracketingNotCentered1_individual`; MdAE = `{r} mdaeBracketingNotCentered1_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError) and [median absolute error](#sec-medianAbsoluteError).

Here is a code example of when the individual predictions bracket but are not centered on the truth for a player, with a high bracketing rate but with an outlier projection that is far away from the truth.
The distribution of projected points versus truth is depicted in @fig-bracketingnotCentered2.

```{r}
predictedValues <- c(40,60,70,80,90,100,110,250)
actualValue <- 74
errors <- predictedValues - actualValue

@@

```{r}
#| label: fig-bracketingnotCentered2
#| fig-cap: "Example Distribution of Projected Fantasy Points that Brackets But is Not Centered on the Truth."
#| fig-alt: "Example Distribution of Projected Fantasy Points that Brackets But is Not Centered on the Truth."
#| code-fold: true

bracketingnotCentered2 <- petersenlab::wisdomOfCrowd(
  predicted = predictedValues,
  actual = actualValue
)

bracketingnotCentered2_bracketingRate <- bracketingnotCentered2[which(row.names(bracketingnotCentered2) == "crowdAveraged"), "bracketingRate"] * 100

maeBracketingnotCentered2_individual <- bracketingnotCentered2[which(row.names(bracketingnotCentered2) == "individual"), "MAE"]
maeBracketingnotCentered2_crowd <- bracketingnotCentered2[which(row.names(bracketingnotCentered2) == "crowdAveraged"), "MAE"]

mdaeBracketingnotCentered2_individual <- bracketingnotCentered2[which(row.names(bracketingnotCentered2) == "individual"), "MdAE"]
meBracketingnotCentered2_crowd <- bracketingnotCentered2[which(row.names(bracketingnotCentered2) == "crowdAveraged"), "ME"]

ggplot2::ggplot(,
  mapping = aes(
    x = predictedValues)
) +
  geom_density(
    fill = "lightgray"
  ) +
  geom_vline(
    xintercept = mean(predictedValues),
    color = "darkgray",
    linewidth = 1.5) +
    geom_vline(
    xintercept = actualValue,
    linetype = "dashed",
    linewidth = 1.5
  ) +
  annotate(
    "segment",
    x = mean(predictedValues),
    xend = actualValue,
    y = 0,
    yend = 0,
    linewidth = 1.5,
    arrow = arrow(
      angle = 20,
      ends = "both",
      type = "closed")
  ) +
  annotate(
    "text",
    x = actualValue + (meBracketingnotCentered2_crowd / 2),
    y = 0.0007,
    label = paste("Mean Error: ", meBracketingnotCentered2_crowd, sep = ""),
    hjust = 0.5 # center-justify
  ) +
  labs(
    x = "Fantasy Points",
    y = "Density",
    title = "Density Plot of Predicted Fantasy Points for a Player",
    subtitle = "Crowd-Averaged Projection (Vertical Gray Line) Vs.\nTruth, i.e., Actual Fantasy Points (Dashed Line)"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title
```

As demonstrated above, when the individual predictions bracket but are not centered on the truth, and the bracketing rate is high (in this case `{r} petersenlab::apa(bracketingnotCentered2_bracketingRate, decimals = 2)`%) but with an outlier projection that is far away from the truth, the crowd-averaged predictions (MAE/MdAE = `{r} maeBracketingnotCentered2_crowd`) are more accurate than the accuracy of the typical individual forecaster (MAE = `{r} maeBracketingnotCentered2_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError), but they can still be less accurate—in terms of [median absolute error](#sec-medianAbsoluteError)—than the individual forecaster who is at the 50th percentile in accuracy ( MdAE = `{r} mdaeBracketingnotCentered2_individual`).

Here is another code example of when the individual predictions bracket but are not centered on the truth for a player, with a low bracketing rate.
The distribution of projected points versus truth is depicted in @fig-bracketingnotCentered3.

```{r}
predictedValues <- c(10,29,29,40,60,70,80,90)
actualValue <- 14
errors <- predictedValues - actualValue

```{r}
#| label: fig-bracketingnotCentered3
#| fig-cap: "Example Distribution of Projected Fantasy Points that Brackets But is Not Centered on the Truth."
#| fig-alt: "Example Distribution of Projected Fantasy Points that Brackets But is Not Centered on the Truth."
#| code-fold: true

bracketingnotCentered3 <- petersenlab::wisdomOfCrowd(
  predicted = predictedValues,
  actual = actualValue
)

bracketingnotCentered3_bracketingRate <- bracketingnotCentered3[which(row.names(bracketingnotCentered3) == "crowdAveraged"), "bracketingRate"] * 100

maeBracketingnotCentered3_individual <- bracketingnotCentered3[which(row.names(bracketingnotCentered3) == "individual"), "MAE"]
maeBracketingnotCentered3_crowd <- bracketingnotCentered3[which(row.names(bracketingnotCentered3) == "crowdAveraged"), "MAE"]

mdaeBracketingnotCentered3_individual <- bracketingnotCentered3[which(row.names(bracketingnotCentered3) == "individual"), "MdAE"]
meBracketingnotCentered3_crowd <- bracketingnotCentered3[which(row.names(bracketingnotCentered3) == "crowdAveraged"), "ME"]

ggplot2::ggplot(,
  mapping = aes(
    x = predictedValues)
) +
  geom_density(
    fill = "lightgray"
  ) +
  geom_vline(
    xintercept = mean(predictedValues),
    color = "darkgray",
    linewidth = 1.5) +
    geom_vline(
    xintercept = actualValue,
    linetype = "dashed",
    linewidth = 1.5
  ) +
  annotate(
    "segment",
    x = mean(predictedValues),
    xend = actualValue,
    y = 0,
    yend = 0,
    linewidth = 1.5,
    arrow = arrow(
      angle = 20,
      ends = "both",
      type = "closed")
  ) +
  annotate(
    "text",
    x = actualValue + (meBracketingnotCentered3_crowd / 2),
    y = 0.0003,
    label = paste("Mean Error: ", meBracketingnotCentered3_crowd, sep = ""),
    hjust = 0.5 # center-justify
  ) +
  labs(
    x = "Fantasy Points",
    y = "Density",
    title = "Density Plot of Predicted Fantasy Points for a Player",
    subtitle = "Crowd-Averaged Projection (Vertical Gray Line) Vs.\nTruth, i.e., Actual Fantasy Points (Dashed Line)"
  ) +
  theme_classic() +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5)) # horizontal y-axis title
```

As demonstrated above, when the individual predictions bracket but are not centered on the truth and the bracketing rate is low (in this case `{r} petersenlab::apa(bracketingnotCentered3_bracketingRate, decimals = 2)`%), the crowd-averaged predictions (MAE/MdAE = `{r} maeBracketingnotCentered3_crowd`) are more accurate than the accuracy of the typical individual forecaster (MAE = `{r} maeBracketingnotCentered3_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError), but they can still be less accurate—in terms of [median absolute error](#sec-medianAbsoluteError)—than the individual forecaster who is at the 50th percentile in accuracy (MdAE = `{r} mdaeBracketingnotCentered3_individual`).

Here is a code example of when the individual predictions do not bracket the truth for a player, and there are no strong outliers.
The distribution of projected points versus truth is depicted in @fig-noBracketing1.

@@

As demonstrated above, when the individual predictions do not bracket the truth (i.e., the bracketing rate is 0%) and there are no strong outliers, the crowd-averaged predictions (MAE/MdAE = `{r} maeNoBracketing1_crowd`) are as accurate as the accuracy of the typical individual forecaster (MAE = `{r} maeNoBracketing1_individual`; MdAE = `{r} mdaeNoBracketing1_individual`) in terms of [mean absolute error](#sec-meanAbsoluteError) and [median absolute error](#sec-medianAbsoluteError).

@@

```{r}
predictedValues <- c(0,65,67,70,80,85,90,95)
actualValue <- 100
errors <- predictedValues - actualValue

@@

In conclusion, as demonstrated from these examples, when the crowd-averaged predictions are centered on the truth, the crowd-averaged predictions are perfectly accurate and are more accurate than the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError) and of most individual forecasters in terms of [median absolute error](#sec-medianAbsoluteError).
When the crowd-averaged predictions bracket but are not centered on the truth, the crowd-averaged predictions are more accurate than the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError).
If the bracketing rate is high and there are no strong outliers, the crowd-averaged predictions will also be more accurate—in terms of [median absolute error](#sec-medianAbsoluteError)—than most individual forecasters; however, if the bracketing rate is low or there are strong outliers, the crowd-averaged predictions can be less accurate than than most individual forecasters.
Outliers could make the crowd-averaged prediction more or less accurate, depending on whether the outlying prediction is closer to or farther away from the truth, compared to the other predictions.
When the crowd-averaged predictions do not bracket the truth, the crowd-averaged predictions are as accurate as the accuracy of the typical individual forecaster in terms of [mean absolute error](#sec-meanAbsoluteError).
When the crowd-averaged predictions do not bracket the truth and there are outliers, the crowd-averaged projections can be less accurate—in terms of [median absolute error](#sec-medianAbsoluteError)—than most individual forecasters (again, depending on whether the outlying prediction is closer to or farther away from the truth, compared to the other predictions).

In sum, going with crowd-averaged projections will always yield predictions that are as or more accurate (in terms of [mean absolute error](#sec-meanAbsoluteError)) than the (collective) accuracy of the individuals' projections.
However, the crowd-averaged projections may be less accurate than any *individual* projection, including the projection of the forecaster who is at the 50th percentile in accuracy.
That is, if there is a low bracketing rate or inaccurate outliers, the crowd-averaged projections can be less accurate than most individual projections.
In general, the higher the bracketing rate and the more closely that the projections are centered on the truth, the more accurate the crowd-averaged projections will be.

There may be multiple ways of handling outlier predictions.
For instance, instead of computing a simple average of the predictions, one could calculate a [weighted average](#sec-descriptiveStatisticsCenter) that weights each prediction according to the historical accuracy of the prediction source.
Alternatively, one could calculate the crowd-averaged prediction using an index of the center of a distribution that is less sensitive to outliers, such as a [median](#sec-descriptiveStatisticsCenter) or robust average such as the [Hodges-Lehmann statistic](#sec-descriptiveStatisticsCenter) (aka pseudomedian).
